---
title: "p8105_jl5934_hw6"
author: "Jesus Luevano"
date: "2023-11-17"
output: github_document
---

```{r, libraries}
library(tidyverse)
library(modelr)
library(purrr)
library(mgcv)
library(caret)
```

# Problem 1

For this problem we will utilize the Washington post data on homicides, to evaluate a generalized logistic regression model to predict resolved cases based on victim age, victim sex, and victim race. 

First we import and clean the data for our purposes. 

```{r, WaPO DF}
WaPo.df <- read_csv("data/homicide-data.csv") %>%
  janitor::clean_names() %>%
  mutate(city_state = 
           paste0(city,", ",state),
         victim_age = as.numeric(victim_age),
         resolved = ifelse(disposition == "Closed by arrest", 1, 0)) %>%
  filter(!(city_state %in% c("Dallas, TX", "Phoenix, AZ", "Kansas City, MO", "Tulsa, AL"))) %>%
  subset(victim_race %in% c("White", "Black"))

```

Next we trial the model on solely Baltimore, MD.

```{r, trial with Balt model}
Balt_model = WaPo.df %>%
  filter(city_state == "Baltimore, MD") |>
  glm(resolved ~ victim_age + victim_sex + victim_race, data = _, family = binomial()) %>%
  broom::tidy() %>%
  filter(term == "victim_sexMale") %>%
  mutate(OR = exp(estimate),
         lower_CI = OR - 1.96*std.error,
         upper_CI = OR + 1.96*std.error)
```

Now we create a model that works for all the cities to output an OR of homcide being resolved based on victim sex of male vs female, and calculate the upper and lower confidence intervals. 

```{r, glm with all cities}
cities = as.vector(unique(WaPo.df$city_state))

OR_sex = function(x) {
  
  model <- WaPo.df %>%
  filter(city_state == x) |>
  glm(resolved ~ victim_age + victim_sex + victim_race, data = _, family = binomial()) %>%
  broom::tidy() %>%
  filter(term == "victim_sexMale") %>%
  mutate(OR = exp(estimate),
         lower_CI = OR - 1.96*std.error,
         upper_CI = OR + 1.96*std.error)
  
  tibble(
    city = x,
    OR = model[["OR"]],
    CI_low = model[["lower_CI"]],
    CI_upper = model[["upper_CI"]]
)
  
}

WaPo_ORs = map(cities, OR_sex) %>%
  bind_rows()
```

Lastly, we plot the the OR of each city, ordered by increasing OR, and error bars for the CI. 

```{r, plot ORs and CIs by city}
ggplot(WaPo_ORs, aes(x = fct_reorder(city, OR), y = OR)) +
  geom_point(stat = "identity") +
  geom_errorbar(aes(ymin = CI_low, ymax = CI_upper)) + 
  theme(axis.text.x = element_text(angle = 45, size = 5, hjust = 1))
```

* We see that there is some range in CI, with smaller ranges in certain cities like Chicago, IL and Detroit, MI. 

# Problem 2

Our goal is to create a linear model from the NOAA weather datasets to predict maximum temperature `tmax` using the variables minimum temperature `tmin` and precipitation `prcp`. From this we will isolate the values of R2 of the model, and the log of the product of the Beta statistics (B1 and B2) after bootstrapping 5000 times. 

First we pull in the dataset. 

```{r, weather dataframe}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2022-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = recode(id, USW00094728 = "CentralPark_NY"),
    tmin = tmin / 10,
    tmax = tmax / 10) |>
  select(name, id, everything())
```

We will set up the bootstrapping function such that it selects a quarter of the total samples each time, with replacement, and creates 5000 new subsets. 

```{r, clean up weather df}
#lm(tmax ~ tmin + prcp, data = weather_df) %>%
 # broom::glance() #%>% #r.squared 
  #broom::tidy() #log(tibble[2,4]*tibble[3,4])

boot_sample = function(df) {
  sample_frac(tbl = df, size = .25, replace = TRUE)
}

boot_straps = 
  tibble(strap_number = 1:5000) |> 
  mutate(
    strap_sample = map(strap_number, \(i) boot_sample(df = weather_df))
  )

boot_straps
```

We will then utilize these bootstrapped subsets and map them over a linear model to predict tmax utilizing tmin and precipitation. I have an additional function to pull in the first cell, to pull out the rsquared from the glance output. For the Betas we will run a separate function to filter out the statistics for tmin and prcp, and mutate a new variable that is the log of their products. 

Both of these will be plotted to show the distribution of them.

```{r, bootstrapping and isolating output of model}
keep_first_cell <- function(tbl) {
  if (nrow(tbl) == 1) {
    return(tbl[1, 1, drop = FALSE])
  } else {
    return(tbl)
  }
}

bootstrap_results = 
  boot_straps %>%
  mutate(
    models = map(strap_sample, \(df) lm(tmax ~ tmin + prcp, data = df)),
    results = map(models, broom::tidy),
    glance = map(models, broom::glance),
    rsquared = map(glance, keep_first_cell)) %>%
  select(-strap_sample, -models, -glance) %>%
  unnest(c(results, rsquared))

bootstrap_results %>%
  ggplot(aes(x = r.squared)) + 
  geom_density()

log_betas = bootstrap_results %>%
  filter(term %in% c("tmin", "prcp")) %>%
  group_by(strap_number) %>%
  summarize(beta_multiple = prod(statistic)) %>%
  mutate(log_beta_multiple = log(beta_multiple))

log_betas %>%
  ggplot(aes(x = log_beta_multiple)) +
  geom_density()
```

* Our distribution of R2 values are primarily located at around 0.93, with a left sided tail nearing zero below 0.8

* Our distribution of log(B1*B2) values are primarily located at 4.2, with a leftward tail extended below zero. 


Next to find the upper and lower quantile (at 0.025 and 0.975 each) we will use the quantile function.

```{r, quantiles}
# Assuming bootstrapped_r2 is a vector of your 5000 bootstrap estimates for R^2
lower_r2 <- quantile(bootstrap_results$r.squared, 0.025)
upper_r2 <- quantile(bootstrap_results$r.squared, 0.975)

# Assuming bootstrapped_log_beta is a vector of your 5000 bootstrap estimates for log(beta0 * beta1), how to deal w/ NaNs? from negative values
lower_log_beta <- quantile(log_betas$log_beta_multiple, 0.025, na.rm = TRUE)
upper_log_beta <- quantile(log_betas$log_beta_multiple, 0.975, na.rm = TRUE)


```

* For R2 the lower quantile at 2.5% is `r print(lower_r2)` 0.8637678, whereas the upper quantile at 97.5% is `r print(upper_r2)` 0.9558145. Thus the 95% CI is between those two values. 

* For log(B1*B2) the lower quantile at 2.5% is `r print(lower_log_beta)` 0.9941103, whereas the upper quantile at 97.5% is `r print(upper_log_beta)` 4.940135. Thus the 95% CI is between those two values. 


# Problem 3

In this problem we will utilize the birthweight dataframe to evaluate different models using the cross-validation functions. 

* First we import that data, clean the names, mutate variables to make new factors of sex, father's race, mother's race, and malformation.  

```{r, birthweight dataframe prep}
birthweight.df <- read_csv("data/birthweight.csv") %>%
  janitor::clean_names() %>%
  mutate(
    sex = as.factor(ifelse(babysex == 1, "male", "female")),
    father_race = as.factor(case_when(
      frace == "1" ~ "White",
      frace == "2" ~ "Black",
      frace == "3" ~ "Asian",
      frace == "4" ~ "Puerto Rican",
      frace == "8" ~ "Other", 
      frace == "9" ~ "Unknown"
    )),
    malformation = as.factor(ifelse(malform == 1, "malformed", "non_malformed")),
    mother_race = as.factor(case_when(
      mrace == "1" ~ "White",
      mrace == "2" ~ "Black",
      mrace == "3" ~ "Asian",
      mrace == "4" ~ "Puerto Rican",
      mrace == "8" ~ "Other"
    ))
  ) %>%
  dplyr::select(-c(babysex, frace, mrace, malform, pnumlbw, pnumsga))
```

I have constructed a model to predict birthweight (`bwt`), that considers socioeconomic demographic factors under that assumption that they are surrogates for health and healthcare access. I chose to model a linear model of `bwt` that relies on family monthly income, mother's weight at delivery, and maternal age. 

This model will be named `fit_mine`. 

```{r, my birthweight model}
fit_mine = lm(bwt ~ fincome + delwt + momage, data = birthweight.df) 

birthweight.df %>%
  modelr::add_residuals(fit_mine) %>%
  modelr::add_predictions(fit_mine) %>%
  ggplot(aes(x = pred, y = resid)) +
  geom_point()
```

* Here we demonstrate a plot of the model's predicted values for birthweight plotted against the residual values from my model. There is wide variability to the residuals, with most from -1000 to +1000, and most of the predicted values being <3500 for birthweight. 


Next we will construct the models described by the homework. 

* The Main model will be named `fit_main`, and predicts birthweight from baby's length at birth and gestational age in weeks. 

* The alternative model will be named `fit_alt`, and predicts birthweight from baby's head circumference, baby's length at birth, the sex of the baby, and interaction terms amongst them all: baby's head circumference * baby's length at birth, baby's head circumference * sex of baby, baby's length at birth * baby's sex, and baby's head circumference * baby's length at birth * baby's sex. 

```{r, main and alternate models for birthweights}
fit_main = lm(bwt ~ blength + gaweeks, data = birthweight.df)
birthweight.df %>%
  modelr::add_residuals(fit_main) %>%
  modelr::add_predictions(fit_main) %>%
  ggplot(aes(x = pred, y = resid)) +
  geom_point()

fit_alt = lm(bwt ~ bhead + blength + sex + bhead*blength + bhead*sex + blength*sex + bhead*blength*sex, data = birthweight.df)
birthweight.df %>%
  modelr::add_residuals(fit_alt) %>%
  modelr::add_predictions(fit_alt) %>%
  ggplot(aes(x = pred, y = resid)) +
  geom_point()
```

* Here we demonstrate two plots of the model's predicted values for birthweight plotted against the residual values from the model. 

  * The first is for the main model using main effects of birth length and gestational age. Compared to my model the residuals are a bit tighter but the spread is still from -1000 to +1000, although there is a greater range of predicted birthweights. 

  * The second is the alternative including the interactions and main effects of head circumference, length at birth, and sex of the baby. Compared to the prior two models, the residual range is a bit tighter now primarily focused in -500 to +500, with still a wide range of predicted values similiar to the main model. 


Next we will set up cross validation of 100 splits. This will include variables for the each of the three above models, mapping the models across the cross-validation dataframe. We will then then calculate the RMSE for each of those models by mapping across them. 

```{r, cross validation}
cv_df = 
  crossv_mc(birthweight.df, 100) 

cv_df = 
  cv_df |> 
  mutate(
    fit_mine  = map(train, \(df) lm(bwt ~ fincome + delwt + momage, data = df)), #fit_mine = lm(bwt ~ fincome + delwt + momage, data = birthweight.df) 
    fit_main  = map(train, \(df) lm(bwt ~ blength + gaweeks, data = df)), #fit_main = lm(bwt ~ blength + gaweeks, data = birthweight.df)
    fit_alt  = map(train, \(df) lm(bwt ~ bhead + blength + sex + bhead*blength + bhead*sex + blength*sex + bhead*blength*sex, data = df))) |>  #fit_alt = lm(bwt ~ bhead + blength + sex + bhead*blength + bhead*sex + blength*sex + bhead*blength*sex, data = birthweight.df)
  mutate(
    rmse_mine = map2_dbl(fit_mine, test, \(mod, df) rmse(model = mod, data = df)),
    rmse_main = map2_dbl(fit_main, test, \(mod, df) rmse(model = mod, data = df)),
    rmse_alt = map2_dbl(fit_alt, test, \(mod, df) rmse(model = mod, data = df)))
```

Lastly we will plot the density of the RMSE of the cross-validations to see which model is optimal. 

```{r, density plot}
cv_df |> 
  select(starts_with("rmse")) |> 
  pivot_longer(
    everything(),
    names_to = "model", 
    values_to = "rmse",
    names_prefix = "rmse_") |> 
  mutate(model = fct_inorder(model)) |> 
  ggplot(aes(x = model, y = rmse)) + geom_violin()
```

* The model with highest RMSE was my own, whereas the alternative that included more variables and interaction terms had the lowest spread of RMSE values. This suggests it is the more accurate amongst them.
