---
title: "p8105_jl5934_hw6"
author: "Jesus Luevano"
date: "2023-11-17"
output: github_document
---

```{r, libraries}
library(tidyverse)
library(modelr)
library(purrr)
library(mgcv)
library(caret)
```

# Problem 1

Create a city_state variable (e.g. “Baltimore, MD”), and a binary variable indicating whether the homicide is solved. Omit cities Dallas, TX; Phoenix, AZ; and Kansas City, MO – these don’t report victim race. Also omit Tulsa, AL – this is a data entry mistake. For this problem, limit your analysis those for whom victim_race is white or black. Be sure that victim_age is numeric.

For the city of Baltimore, MD, use the glm function to fit a logistic regression with resolved vs unresolved as the outcome and victim age, sex and race as predictors. Save the output of glm as an R object; apply the broom::tidy to this object; and obtain the estimate and confidence interval of the adjusted odds ratio for solving homicides comparing male victims to female victims keeping all other variables fixed.

Now run glm for each of the cities in your dataset, and extract the adjusted odds ratio (and CI) for solving homicides comparing male victims to female victims. Do this within a “tidy” pipeline, making use of purrr::map, list columns, and unnest as necessary to create a dataframe with estimated ORs and CIs for each city.

Create a plot that shows the estimated ORs and CIs for each city. Organize cities according to estimated OR, and comment on the plot.



```{r, WaPO DF}
WaPo.df <- read_csv("data/homicide-data.csv") %>%
  janitor::clean_names() %>%
  mutate(city_state = 
           paste0(city,", ",state),
         victim_age = as.numeric(victim_age),
         resolved = ifelse(disposition == "Closed by arrest", 1, 0)) %>%
  filter(!(city_state %in% c("Dallas, TX", "Phoenix, AZ", "Kansas City, MO", "Tulsa, AL"))) %>%
  subset(victim_race %in% c("White", "Black"))

```

```{r, trial with Balt model}
Balt_model = WaPo.df %>%
  filter(city_state == "Baltimore, MD") |>
  glm(resolved ~ victim_age + victim_sex + victim_race, data = _, family = binomial()) %>%
  broom::tidy() %>%
  filter(term == "victim_sexMale") 
  #mutate(OR = exp(estimate),
        # upper_CI = )
```



# Problem 2

Our goal is to create a linear model from the NOAA weather datasets to predict maximum temperature `tmax` using the variables minimum temperature `tmin` and precipitation `prcp`. From this we will isolate the values of R2 of the model, and the log of the product of the Beta statistics (B1 and B2) after bootstrapping 5000 times. 

First we pull in the dataset. 

```{r, weather dataframe}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2022-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = recode(id, USW00094728 = "CentralPark_NY"),
    tmin = tmin / 10,
    tmax = tmax / 10) |>
  select(name, id, everything())
```

We will set up the bootstrapping function such that it selects a quarter of the total samples each time, with replacement, and creates 5000 new subsets. 

```{r, clean up weather df}
#lm(tmax ~ tmin + prcp, data = weather_df) %>%
 # broom::glance() #%>% #r.squared 
  #broom::tidy() #log(tibble[2,4]*tibble[3,4])

boot_sample = function(df) {
  sample_frac(tbl = df, size = .25, replace = TRUE)
}

boot_straps = 
  tibble(strap_number = 1:5000) |> 
  mutate(
    strap_sample = map(strap_number, \(i) boot_sample(df = weather_df))
  )

boot_straps
```

We will then utilize these bootstrapped subsets and map them over a linear model to predict tmax utilizing tmin and precipitation. I have an additional function to pull in the first cell, to pull out the rsquared from the glance output. For the Betas we will run a separate function to filter out the statistics for tmin and prcp, and mutate a new variable that is the log of their products. 

Both of these will be plotted to show the distribution of them.

```{r, bootstrapping and isolating output of model}
keep_first_cell <- function(tbl) {
  if (nrow(tbl) == 1) {
    return(tbl[1, 1, drop = FALSE])
  } else {
    return(tbl)
  }
}

bootstrap_results = 
  boot_straps %>%
  mutate(
    models = map(strap_sample, \(df) lm(tmax ~ tmin + prcp, data = df)),
    results = map(models, broom::tidy),
    glance = map(models, broom::glance),
    rsquared = map(glance, keep_first_cell)) %>%
  select(-strap_sample, -models, -glance) %>%
  unnest(c(results, rsquared))

bootstrap_results %>%
  ggplot(aes(x = r.squared)) + 
  geom_density()

log_betas = bootstrap_results %>%
  filter(term %in% c("tmin", "prcp")) %>%
  group_by(strap_number) %>%
  summarize(beta_multiple = prod(statistic)) %>%
  mutate(log_beta_multiple = log(beta_multiple))

log_betas %>%
  ggplot(aes(x = log_beta_multiple)) +
  geom_density()
```

* Our distribution of R2 values are primarily located at around 0.93, with a left sided tail nearing zero below 0.8

* Our distribution of log(B1*B2) values are primarily located at 4.2, with a leftward tail extended below zero. 


Next to find the upper and lower quantile (at 0.025 and 0.975 each) we will use the quantile function.

```{r, quantiles}
# Assuming bootstrapped_r2 is a vector of your 5000 bootstrap estimates for R^2
lower_r2 <- quantile(bootstrap_results$r.squared, 0.025)
upper_r2 <- quantile(bootstrap_results$r.squared, 0.975)

# Assuming bootstrapped_log_beta is a vector of your 5000 bootstrap estimates for log(beta0 * beta1), how to deal w/ NaNs? from negative values
lower_log_beta <- quantile(log_betas$log_beta_multiple, 0.025, na.rm = TRUE)
upper_log_beta <- quantile(log_betas$log_beta_multiple, 0.975, na.rm = TRUE)


```

* For R2 the lower quantile at 2.5% is `r print(lower_r2)` 0.8637678, whereas the upper quantile at 97.5% is `r print(upper_r2)` 0.9558145. Thus the 95% CI is between those two values. 

* For log(B1*B2) the lower quantile at 2.5% is `r print(lower_log_beta)` 0.9941103, whereas the upper quantile at 97.5% is `r print(upper_log_beta)` 4.940135. Thus the 95% CI is between those two values. 


# Problem 3

In this problem we will utilize the birthweight dataframe to evaluate different models using the cross-validation functions. 

* First we import that data, clean the names, mutate variables to make new factors of sex, father's race, mother's race, and malformation.  

```{r, birthweight dataframe prep}
birthweight.df <- read_csv("data/birthweight.csv") %>%
  janitor::clean_names() %>%
  mutate(
    sex = as.factor(ifelse(babysex == 1, "male", "female")),
    father_race = as.factor(case_when(
      frace == "1" ~ "White",
      frace == "2" ~ "Black",
      frace == "3" ~ "Asian",
      frace == "4" ~ "Puerto Rican",
      frace == "8" ~ "Other", 
      frace == "9" ~ "Unknown"
    )),
    malformation = as.factor(ifelse(malform == 1, "malformed", "non_malformed")),
    mother_race = as.factor(case_when(
      mrace == "1" ~ "White",
      mrace == "2" ~ "Black",
      mrace == "3" ~ "Asian",
      mrace == "4" ~ "Puerto Rican",
      mrace == "8" ~ "Other"
    ))
  ) %>%
  dplyr::select(-c(babysex, frace, mrace, malform, pnumlbw, pnumsga))
```

I have constructed a model to predict birthweight (`bwt`), that considers socioeconomic demographic factors under that assumption that they are surrogates for health and healthcare access. I chose to model a linear model of `bwt` that relies on family monthly income, mother's weight at delivery, and maternal age. 

This model will be named `fit_mine`. 

```{r, my birthweight model}
fit_mine = lm(bwt ~ fincome + delwt + momage, data = birthweight.df) 

birthweight.df %>%
  modelr::add_residuals(fit_mine) %>%
  modelr::add_predictions(fit_mine) %>%
  ggplot(aes(x = pred, y = resid)) +
  geom_point()
```

* Here we demonstrate a plot of the model's predicted values for birthweight plotted against the residual values from my model.


Next we will construct the models described by the homework. 

* The Main model will be named `fit_main`, and predicts birthweight from baby's length at birth and gestational age in weeks. 

* The alternative model will be named `fit_alt`, and predicts birthweight from baby's head circumference, baby's length at birth, the sex of the baby, and interaction terms amongst them all: baby's head circumference * baby's length at birth, baby's head circumference * sex of baby, baby's length at birth * baby's sex, and baby's head circumference * baby's length at birth * baby's sex. 

```{r, main and alternate models for birthweights}
fit_main = lm(bwt ~ blength + gaweeks, data = birthweight.df)
birthweight.df %>%
  modelr::add_residuals(fit_main) %>%
  modelr::add_predictions(fit_main) %>%
  ggplot(aes(x = pred, y = resid)) +
  geom_point()

fit_alt = lm(bwt ~ bhead + blength + sex + bhead*blength + bhead*sex + blength*sex + bhead*blength*sex, data = birthweight.df)
birthweight.df %>%
  modelr::add_residuals(fit_alt) %>%
  modelr::add_predictions(fit_alt) %>%
  ggplot(aes(x = pred, y = resid)) +
  geom_point()
```

* Here we demonstrate two plots of the model's predicted values for birthweight plotted against the residual values from the model. The first is for the main model using main effects of birth length and gestatational age. The second is the alternative including the interactions and main effects of head circumference, length at birth, and sex of the baby.


Next we will set up cross validation of 100 splits. This will include variables for the each of the three above models, mapping the models across the cross-validation dataframe. We will then then calculate the RMSE for each of those models by mapping across them. 

```{r, cross validation}
cv_df = 
  crossv_mc(birthweight.df, 100) 

cv_df = 
  cv_df |> 
  mutate(
    fit_mine  = map(train, \(df) lm(bwt ~ fincome + delwt + momage, data = df)), #fit_mine = lm(bwt ~ fincome + delwt + momage, data = birthweight.df) 
    fit_main  = map(train, \(df) lm(bwt ~ blength + gaweeks, data = df)), #fit_main = lm(bwt ~ blength + gaweeks, data = birthweight.df)
    fit_alt  = map(train, \(df) lm(bwt ~ bhead + blength + sex + bhead*blength + bhead*sex + blength*sex + bhead*blength*sex, data = df))) |>  #fit_alt = lm(bwt ~ bhead + blength + sex + bhead*blength + bhead*sex + blength*sex + bhead*blength*sex, data = birthweight.df)
  mutate(
    rmse_mine = map2_dbl(fit_mine, test, \(mod, df) rmse(model = mod, data = df)),
    rmse_main = map2_dbl(fit_main, test, \(mod, df) rmse(model = mod, data = df)),
    rmse_alt = map2_dbl(fit_alt, test, \(mod, df) rmse(model = mod, data = df)))
```

Lastly we will plot the density of the RMSE of the cross-validations to see which model is optimal. 

```{r, density plot}
cv_df |> 
  select(starts_with("rmse")) |> 
  pivot_longer(
    everything(),
    names_to = "model", 
    values_to = "rmse",
    names_prefix = "rmse_") |> 
  mutate(model = fct_inorder(model)) |> 
  ggplot(aes(x = model, y = rmse)) + geom_violin()
```

* The model with highest RMSE was my own, whereas the alternative that included more variables and interaction terms had the lowest spread of RMSE values. This suggests it is the more accurate amongst them.
